---
title: "Building an app from files and YAML"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The [data model article](data-model.html) shows how to assemble an
`ExploratorySummarizedExperimentList` (ESEList) in R. That is the right approach
when your results are already in memory. But a bioinformatics pipeline typically
finishes with a *directory of files* - expression matrices, a sample sheet,
feature annotation, a contrasts table, differential-results tables and maybe
enrichment output. `shinyngs` can read that file complement directly, so you can
turn a pipeline's outputs into a running app without writing any bespoke R.

There are two file-driven routes:

* **`eselist_from_yaml()`** - an R function that reads a YAML manifest pointing at
  your files and returns an ESEList you can pass to `prepare_app()`.
* **`exec/make_app_from_files.R`** - a command-line script that takes the same
  kinds of files as flags and writes a ready-to-run app bundle (`app.R` +
  `data.rds`), optionally deploying it to shinyapps.io.

All code on this page is shown but not executed at build time (it needs your
files and, in places, a browser), so treat the snippets as templates to adapt.

## The YAML route

`eselist_from_yaml("my.yaml")` reads a manifest that describes the study and
points at the data files. It builds exactly the same object you would build by
hand, so it slots straight into `prepare_app()`:

```{r yaml-call}
library(shinyngs)
eselist <- eselist_from_yaml("my.yaml")
app <- prepare_app("rnaseq", eselist)
shiny::shinyApp(app$ui, app$server)
```

A representative manifest:

```yaml
title: My RNA-seq experiment
author: Joe Bloggs
report: report.md
ensembl_species: hsapiens
group_vars:
  - Group
  - Replicate
default_groupvar: Group
experiments:
  Gene:
    coldata:
      file: samplesheet.csv
      id: sample
    annotation:
      file: gene_meta.csv
      id: gene_id
      label: gene_name
    expression_matrices:
      Raw:
        file: raw_counts.csv
        measure: counts
      Normalised:
        file: normalised_counts.csv
        measure: Counts per million
    read_reports:
      read_attrition: read_attrition.csv
contrasts:
  comparisons:
    0:
      - Group
      - control
      - treated
stats:
  Gene:
    Normalised:
      pvals: pvals.csv
      qvals: qvals.csv
```

Mapping this back to the [data model](data-model.html): each entry under
`experiments` becomes one `ExploratorySummarizedExperiment` (so you can describe
transcript- and gene-level experiments side by side); `coldata`, `annotation`
and `expression_matrices` populate the ESE's `colData`, annotation and assays;
`id`/`label` set the `idfield`/`labelfield`; `contrasts` and `stats` populate
the study contrasts and the per-ESE `contrast_stats`.

Useful optional top-level keys:

* **`ensembl_species`** (e.g. `hsapiens`, `mmusculus`) - when set, and the
  annotation supplies `chromosome_name`/`start_position`/`end_position`, enables
  the igv.js gene-model view in the `gene` module.
* **`static_pdf`** - switch static plot downloads to PDF.
* **`url_roots`** - add link-outs for identifier columns.
* **`gene_set_id_type`** and **`gene_sets`** (a named list of `.gmt` paths, one
  per collection) - a YAML-driven alternative to supplying gene sets via the R
  constructor.

This is how the `zhangneurons` example dataset was generated.

## The command-line route

`exec/make_app_from_files.R` builds an app bundle from flat files without
writing any R. It is the path used by pipelines such as
[nf-core/differentialabundance](https://nf-co.re/differentialabundance), which
calls it to produce the Shiny report at the end of a run.

A minimal invocation:

```sh
make_app_from_files.R \
  --assay_files raw.tsv,normalised_counts.tsv \
  --assay_names Raw,Normalised \
  --sample_metadata samplesheet.csv \
  --feature_metadata gene_meta.tsv \
  --contrast_file contrasts.csv \
  --differential_results treatment-saline-drug.deseq2.results.tsv \
  --contrast_stats_assay 2 \
  --fold_change_scale log2 \
  --output_directory app
```

This writes `app.R` and a serialised `data.rds` into `app/`. The flags map onto
the same data model:

* `--assay_files` / `--assay_names` - the expression matrices and their display
  names.
* `--sample_metadata` / `--sample_id_col` - the sample sheet and its identifier
  column.
* `--feature_metadata` / `--feature_id_col` / `--feature_name_col` - feature
  annotation and which columns are the `idfield` and `labelfield`.
* `--contrast_file` - a CSV whose first three columns are the contrast
  `variable`, `reference` and `target`, one row per contrast.
* `--differential_results` - one differential table per contrast row, read for
  fold changes and p-values (columns named by `--fold_change_column`,
  `--pval_column`, `--qval_column`).
* `--contrast_stats_assay` - the 1-based index of the assay to associate with
  those statistics (usually the normalised matrix).
* `--fold_change_scale` - `log2`, `linear` or `auto`; `auto` infers the scale
  from the column name and value distribution and errors if the signals
  disagree. Whenever the resolved scale is `log2`, the fold changes stored in
  the app are converted to linear - shinyngs always works with fold changes on
  a linear scale internally, so a `log2FoldChange`-named column (or one whose
  values are distributed like log2 fold changes) will not appear in the app
  as it does in the input file. Pass `--fold_change_scale linear` if you want
  the input file's values used as-is.

For the complete flag reference, including grouping variables, gene-model
support (`--ensembl_species`) and shinyapps.io deployment (`--deploy_app`), see
the [CLI reference](cli.html).

### Running the generated bundle

The output directory is a standard Shiny app. Run it locally:

```{r run-bundle}
library(shinyngs)
shiny::runApp("app")
```

Or run it via Docker without installing anything. The
[Biocontainers image](https://quay.io/repository/biocontainers/r-shinyngs)
ships `shinyngs` and Shiny ready to serve the generated `app.R`/`data.rds`.
There is no `latest` tag, so pick the most recent from the
[tag list](https://quay.io/repository/biocontainers/r-shinyngs?tab=tags) and
mount your output directory:

```sh
docker run --rm -p 3838:3838 \
  -v "$(pwd)/app":/data -w /data \
  quay.io/biocontainers/r-shinyngs:<latest-tag> \
  Rscript -e "shiny::runApp('/data', host = '0.0.0.0', port = 3838)"
```

Then browse to `http://localhost:3838`.

## Wiring in enrichment results

`make_app_from_files.R` can also attach GSEA (or `roast`) results from the
command line. `--enrichment_filename_template` describes where result files live,
with `{contrast_name}` and `{geneset_type}` substituted per contrast and gene-set
type, plus an optional `{target|reference}` token that points at the split
up/down GSEA files:

```sh
make_app_from_files.R \
  ... \
  --contrast_file contrasts.csv \
  --enrichment_gene_sets my_gene_sets.gmt \
  --enrichment_filename_template 'gsea_{contrast_name}_{geneset_type}_for_{target|reference}.tsv'
```

Notes:

* `--enrichment_gene_sets` takes a comma-separated list of the GMT files used in
  the analysis; `--enrichment_gene_type_id` names the identifier type they use
  (e.g. `gene_name` or an Entrez ID).
* `--enrichment_skip_missing` tolerates contrasts or gene-set types with no
  result file; those simply show "No enrichment results for this contrast" in
  the app.
* For a tool that is not auto-detected, name the columns to filter on with
  `--enrichment_pval_column`, `--enrichment_fdr_column` and
  `--enrichment_direction_column`. Auto-detection of `roast`/GSEA formats is
  described in the [data model article](data-model.html#tool-formats-and-auto-detection).

## Other command-line tools

The `exec/` directory holds a few more scripts that reuse `shinyngs`' plotting
and validation code outside the Shiny app. Each takes `--help`:

* `exploratory_plots.R` - static exploratory plots (PCA, boxplots, densities)
  from assay files and sample metadata.
* `differential_plots.R` - static differential plots (volcano) from
  differential-results files.
* `validate_fom_components.R` - runs `shinyngs`' consistency checks (matrices vs.
  sample/feature annotation vs. contrasts) on a file complement *without*
  building an app, so a feature/observation-matrix pipeline can fail fast on
  inconsistent inputs.

See the [CLI reference](cli.html) for the full argument set of each.
